---
title: "Reproducing analysis of Karthaus et al Science 2020"
author: "Lian Chee Foong"
date: "2023-05-07"
output: html_document
---

#### Load Karthaus raw count data ####

```{r}

# Load libraries
library(Seurat)
library(tidyverse)
library(stringr)
library(rliger)
library(SeuratWrappers)

# load Karthaus CSawyers
CSawyer <- Read10X_h5("E:/R_paper_Karthaus_CSawyer_Sci2020_mouse_castrated_scRNA/Data/GSE146811_mmProstate10x_timecourse_rawCount.h5")
GSE146811_mmProstate10x_full_sample_final_tsv <- read_tsv("E:/R_paper_Karthaus_CSawyer_Sci2020_mouse_castrated_scRNA/Data/GSE146811_mmProstate10x_full_sample_final.tsv.gz",
                                                          trim_ws = TRUE)
GSE146811_mmProstate10x_full_sample_final_tsv <- data.frame(GSE146811_mmProstate10x_full_sample_final_tsv)
rownames(GSE146811_mmProstate10x_full_sample_final_tsv) <- GSE146811_mmProstate10x_full_sample_final_tsv$sampleID
CSawyer <- CreateSeuratObject(counts = CSawyer,
                              meta.data = GSE146811_mmProstate10x_full_sample_final_tsv)
table(CSawyer$orig.ident)
table(CSawyer$batchID)

# add sampleInfo in metadata
metadata <- CSawyer@meta.data
metadata$sampleType <- ifelse(metadata$orig.ident == "T00", "Intact",
                              ifelse(metadata$orig.ident %in% c("T01", "T02", "T03", "T04"),
                                     "Cast", "Regen"))
metadata$sampleInfo <- str_sub(metadata$batchID, 0, -3)
CSawyer@meta.data <- metadata
rm(GSE146811_mmProstate10x_full_sample_final_tsv, metadata)
table(CSawyer$predType)
table(CSawyer$highLevelPred)
```

#### 1. Clustering using LIGER for NMF and graph-based approach (for CSawyer_intact)  ####

```{r}

# Subset Intact samples only 
Idents(CSawyer) <- CSawyer$batchID
CSawyer_intact <- subset(CSawyer, idents = c("T00_intact_1",  "T00_intact_2" ) , invert = F)

CSawyer_intact
table(CSawyer_intact$orig.ident)
table(CSawyer_intact$batchID)

# Compute percent mito ratio
CSawyer_intact$mitoRatio <- PercentageFeatureSet(object = CSawyer_intact, pattern = "^mt-")
CSawyer_intact$mitoRatio <- CSawyer_intact@meta.data$mitoRatio / 100


# filter
filtered_CSawyer_intact <- subset(
  x = CSawyer_intact,
  subset =  nFeature_RNA > 500 &
    nCount_RNA > 1000 &
    mitoRatio < 0.15)

filtered_CSawyer_intact <- NormalizeData(filtered_CSawyer_intact, normalization.method = "LogNormalize", scale.factor = 10000)
filtered_CSawyer_intact <- FindVariableFeatures(filtered_CSawyer_intact, selection.method = "vst", nfeatures = 2000, verbose = T)
filtered_CSawyer_intact <- ScaleData(filtered_CSawyer_intact, split.by = "orig.ident", do.center = FALSE)
filtered_CSawyer_intact <- RunOptimizeALS(filtered_CSawyer_intact, k = 20, lambda = 5, split.by = "orig.ident")
filtered_CSawyer_intact <- RunQuantileNorm(filtered_CSawyer_intact, split.by = "orig.ident")
filtered_CSawyer_intact <- FindNeighbors(filtered_CSawyer_intact, reduction = "iNMF_raw", dims = 1:20)
filtered_CSawyer_intact <- FindClusters(filtered_CSawyer_intact, resolution = c(0.5, 0.8,1,3))
filtered_CSawyer_intact <- RunUMAP(filtered_CSawyer_intact, dims = 1:ncol(filtered_CSawyer_intact[["iNMF_raw"]]), reduction = "iNMF_raw")
Idents(filtered_CSawyer_intact) <- filtered_CSawyer_intact$RNA_snn_res.0.5
DimPlot(filtered_CSawyer_intact, split.by = "orig.ident", label = F)

cols <- rev(rainbow(7)[-7])
gene <- c("Krt5","Krt14","Trp63","Krt8","Krt18","Cd24a")
for (j in 1:length(gene)){
  p <- FeaturePlot(filtered_CSawyer_intact, features = gene[j], split.by = "orig.ident", cols = cols)
  assign(paste0("UMAP_", gene[j]), p)
}
gene <- c("Epcam","Pax2","Pate4","Calml3","Ptprc", "Cdh1")
for (j in 1:length(gene)){
  p <- FeaturePlot(filtered_CSawyer_intact, features = gene[j],split.by = "orig.ident", cols = cols) 
  assign(paste0("UMAP_", gene[j]), p)
}
p1 <- plot_grid(UMAP_Epcam, UMAP_Cdh1,  UMAP_Ptprc,ncol = 1 )
p2 <- plot_grid(UMAP_Trp63, UMAP_Krt5, UMAP_Krt14,  ncol = 1)
p3 <- plot_grid(UMAP_Cd24a, UMAP_Krt8, UMAP_Krt18,  ncol = 1)
p4 <- plot_grid(UMAP_Pax2, UMAP_Pate4, UMAP_Calml3, ncol = 1)
plot_grid(p1,p4, ncol = 2)
plot_grid(p2, p3,  ncol = 2)
```

```{r}
# # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # #
#### classify high-level cell types (epithelial, immune, stromal) ####
# # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # # #
Idents(filtered_CSawyer_intact) <- filtered_CSawyer_intact$RNA_snn_res.0.5
AvgExpr <- AverageExpression(filtered_CSawyer_intact, features = c("Epcam","Ptprc"))
AvgExpr <- AvgExpr[[1]]
AvgExpr <- as.data.frame(AvgExpr)
AvgExpr <- t(AvgExpr)

Idents(filtered_CSawyer_intact) <- "RNA_snn_res.0.5"
filtered_CSawyer_intact <-  RenameIdents(object = filtered_CSawyer_intact,
                                         "0" = "Epithelial",
                                         "1" = "Epithelial",
                                         "2" = "Stromal",
                                         "3" = "Epithelial",
                                         "4" = "Epithelial",
                                         "5" = "Stromal",
                                         "6" = "Stromal",
                                         "7" = "Stromal",
                                         "8" = "Epithelial",
                                         "9" = "Immune",
                                         "10" = "Epithelial",
                                         "11" = "Immune",
                                         "12" = "Epithelial",
                                         "13" = "Immune",
                                         "14" = "Immune",
                                         "15" = "Epithelial",
                                         "16" = "Immune",
                                         "17" = "Epithelial",
                                         "18" = "Stromal",
                                         "19" = "Epithelial",
                                         "20" = "Stromal",
                                         "21" = "Stromal"
)
filtered_CSawyer_intact$highlevelCelltype <- Idents(filtered_CSawyer_intact)
filtered_CSawyer_intact$highLevelPred <- factor(filtered_CSawyer_intact$highLevelPred, levels = c("Epi","Str","Imm"))
p1 <- DimPlot(filtered_CSawyer_intact, group.by = "highlevelCelltype", label = F)
p2 <- DimPlot(filtered_CSawyer_intact, group.by = "highLevelPred", label = F)
plot_grid(p1,p2, ncol = 2)

# save(filtered_CSawyer_intact, file = "filtered_CSawyer_intact_notbeingsubset.RData")

```

#### 2a. Subset to Epithelial cells only (filtered_CSawyer_intact) ####

```{r}

Idents(filtered_CSawyer_intact) <- filtered_CSawyer_intact$highlevelCelltype
filtered_CSawyer_intact_Epi <- subset(filtered_CSawyer_intact, idents = "Epithelial")
filtered_CSawyer_intact_Epi <- CreateSeuratObject(counts = filtered_CSawyer_intact_Epi@assays$RNA@counts,
                                                  meta.data = filtered_CSawyer_intact_Epi@meta.data %>% 
                                                    select(batchID,highLevelPred,highLevelPredAmbig,predType, predTypeAmbig,sampleInfo,highlevelCelltype))

filtered_CSawyer_intact_Epi <- NormalizeData(filtered_CSawyer_intact_Epi, normalization.method = "LogNormalize", scale.factor = 10000)
filtered_CSawyer_intact_Epi <- FindVariableFeatures(filtered_CSawyer_intact_Epi, selection.method = "vst", nfeatures = 2000, verbose = T)
filtered_CSawyer_intact_Epi <- ScaleData(filtered_CSawyer_intact_Epi,  do.center = FALSE)
filtered_CSawyer_intact_Epi <- RunOptimizeALS(filtered_CSawyer_intact_Epi, k = 21, lambda = 5)
filtered_CSawyer_intact_Epi <- RunQuantileNorm(filtered_CSawyer_intact_Epi)
filtered_CSawyer_intact_Epi <- FindNeighbors(filtered_CSawyer_intact_Epi, reduction = "iNMF_raw", dims = 1:20)
filtered_CSawyer_intact_Epi <- FindClusters(filtered_CSawyer_intact_Epi, resolution = c(0.5, 0.8,1,3,5))
filtered_CSawyer_intact_Epi <- RunUMAP(filtered_CSawyer_intact_Epi, dims = 1:ncol(filtered_CSawyer_intact_Epi[["iNMF_raw"]]), reduction = "iNMF_raw")
Idents(filtered_CSawyer_intact_Epi) <- filtered_CSawyer_intact_Epi$RNA_snn_res.0.5


# Check cell type
# refer to Karthaus et al. Science 2020 (CSawyers) mouse clustering
# basal, Epcam, Krt5, Krt14 Trp63;
# seminal vesicle basal, Epcam, Pax2, Krt5, Krt14 Trp63, Calml3;
# luminal 1, Epcam, CD24a, Krt8, Krt18, Nkx3.1, Pbsn high;
# luminal 2, Epcam, CD24a, Krt8, Krt18, Psca, Krt4, Tacst2,Ly6a;
# luminal 3 (ionocyte), Epcam, CD24a, Krt8, Krt18, Foxi1, Atp6v1g3, Atp6b1b;
# seminal vesicle luminal, Epcam, Pax2, Krt8, Krt18, Pate4;
# seminal vesicle ionocyte, Epcam, Pax2, Foxi1.

gene_epi <- c("Epcam", "Krt5", "Krt14", "Trp63", "Calml3",
              "Cd24a", "Krt8", "Krt18", "Nkx3-1", "Pbsn",
              "Psca", "Krt4", "Tacstd2","Ly6a",
              "Foxi1", "Atp6v1g3", "Atp6v1b1", 
              "Pax2","Pate4")
Idents(filtered_CSawyer_intact_Epi) <- filtered_CSawyer_intact_Epi$RNA_snn_res.0.5
AvgExpr <- AverageExpression(filtered_CSawyer_intact_Epi, features = gene_epi)
AvgExpr <- AvgExpr[[1]]
AvgExpr <- as.data.frame(AvgExpr)
AvgExpr <- t(AvgExpr)
AvgExpr <- round(AvgExpr, 3)

Idents(filtered_CSawyer_intact_Epi) <- filtered_CSawyer_intact_Epi$RNA_snn_res.0.5
VlnPlot(filtered_CSawyer_intact_Epi, features = gene_epi, stack = T, flip = T)

Idents(filtered_CSawyer_intact_Epi) <- "RNA_snn_res.0.5"
filtered_CSawyer_intact_Epi <-  RenameIdents(object = filtered_CSawyer_intact_Epi,
                                             "0" = "Basal",
                                             "1" = "Luminal 1",
                                             "2" = "Basal",
                                             "3" = "Basal",
                                             "4" = "Luminal 1",
                                             "5" = "Luminal 1",
                                             "6" = "Basal",
                                             "7" = "Seminal vesicle luminal",
                                             "8" = "Basal",
                                             "9" = "Seminal vesicle luminal",
                                             "10" = "Luminal 3",
                                             "11" = "Luminal 2",
                                             "12" = "Basal",
                                             "13" = "Luminal 1",
                                             "14" = "Luminal 1",
                                             "15" = "Luminal 3"
)
filtered_CSawyer_intact_Epi$luminal_celltype <- Idents(filtered_CSawyer_intact_Epi)
p1 <- DimPlot(filtered_CSawyer_intact_Epi, group.by = "luminal_celltype")
p2 <- DimPlot(filtered_CSawyer_intact_Epi, group.by = "RNA_snn_res.0.5")
plot_grid(p1, p2, ncol = 2)
# save(filtered_CSawyer_intact_Epi, file = "filtered_CSawyer_intact_Epi_XGBoost.RData")
```

#### xgboost for Intact CSawyers (included basal) ####

```{r}
#install.packages("xgboost")
library(xgboost)
library(tidymodels)

####  train using basal and Luminal subtypes and SV luminal #### 
set.seed(123)
data<- filtered_CSawyer_intact_Epi@assays$RNA@scale.data
data<- data[rownames(data) %in% c(gene_epi),]
data<- t(data) %>% as.data.frame()
data$cell_type <- filtered_CSawyer_intact_Epi$luminal_celltype
data$cell_type<- factor(data$cell_type)
data$cell_barcode<- rownames(data)

data_split <- initial_split(data, strata = "cell_type")
data_train <- training(data_split)
data_test <- testing(data_split)

# 10 fold cross validation
data_fold <- vfold_cv(data_train, v = 10)

# Define the recipe for pre-processing the data
bt_recipe <- 
  recipe(formula = cell_type ~ ., data = data_train) %>%
  update_role(cell_barcode, new_role = "ID") %>%
  step_zv(all_predictors())

## Define the model specification and tune hyperparameters
## feature importance sore to TRUE, tune mtry and number of trees
bt_spec <- 
  boost_tree(mtry = tune(), trees = tune()) %>%
  set_engine("xgboost", importance = TRUE) %>%
  set_mode("classification")

bt_workflow <- workflow() %>% 
  add_recipe(bt_recipe) %>% 
  add_model(bt_spec)

bt_grid<- grid_regular(mtry(range= c(1, 2000)), trees(), 
                       levels = 3)

library(doParallel)
library(recipes)
doParallel::registerDoParallel()

# # save_pred = TRUE for later ROC curve
tune_res <- tune_grid(
  bt_workflow,
  resamples = data_fold, 
  grid = bt_grid,
  control = control_grid(save_pred = TRUE)
)

# check which penalty is the best 
autoplot(tune_res)
best_penalty <- select_best(tune_res, metric = "accuracy")
best_penalty

# Fit the model on the training data and evaluate performance on the testing data
bt_final <- finalize_workflow(bt_workflow, best_penalty)
bt_final_fit_Epi <- fit(bt_final, data = data_train)

# confusion matrix, perfect classification! 
predict(bt_final_fit_Epi, new_data = data_test) %>%
  bind_cols(data_test %>% select(cell_type)) %>%
  conf_mat(truth = cell_type, estimate = .pred_class)

# save(bt_final_fit_Epi, file = "RData/XGBoost_CSawyer_T00int_Epi_withbasalsubsets.RData")
```

#### 2b. Subset to Immune only (filtered_CSawyer_intact) ####

```{r}

Idents(filtered_CSawyer_intact) <- filtered_CSawyer_intact$highlevelCelltype
filtered_CSawyer_intact_Imm <- subset(filtered_CSawyer_intact, idents = "Immune")
filtered_CSawyer_intact_Imm <- CreateSeuratObject(counts = filtered_CSawyer_intact_Imm@assays$RNA@counts,
                                                  meta.data = filtered_CSawyer_intact_Imm@meta.data %>% 
                                                    select(batchID,highLevelPred,highLevelPredAmbig,predType, predTypeAmbig,sampleInfo,highlevelCelltype))

filtered_CSawyer_intact_Imm <- NormalizeData(filtered_CSawyer_intact_Imm, normalization.method = "LogNormalize", scale.factor = 10000)
filtered_CSawyer_intact_Imm <- FindVariableFeatures(filtered_CSawyer_intact_Imm, selection.method = "vst", nfeatures = 2000, verbose = T)
filtered_CSawyer_intact_Imm <- ScaleData(filtered_CSawyer_intact_Imm,  do.center = FALSE)
filtered_CSawyer_intact_Imm <- RunOptimizeALS(filtered_CSawyer_intact_Imm, k = 21, lambda = 5)
filtered_CSawyer_intact_Imm <- RunQuantileNorm(filtered_CSawyer_intact_Imm)
filtered_CSawyer_intact_Imm <- FindNeighbors(filtered_CSawyer_intact_Imm, reduction = "iNMF_raw", dims = 1:20)
filtered_CSawyer_intact_Imm <- FindClusters(filtered_CSawyer_intact_Imm, resolution = c(0.5, 0.8,1,3,5))
filtered_CSawyer_intact_Imm <- RunUMAP(filtered_CSawyer_intact_Imm, dims = 1:ncol(filtered_CSawyer_intact_Imm[["iNMF_raw"]]), reduction = "iNMF_raw")
Idents(filtered_CSawyer_intact_Imm) <- filtered_CSawyer_intact_Imm$RNA_snn_res.0.5

DimPlot(filtered_CSawyer_intact_Imm, group.by = "RNA_snn_res.0.5", label = T)

# Check cell type
# refer to Karthaus et al. Science 2020 (CSawyers) mouse clustering
# B-cells, CD19, Ms4a1: 
# T-cells, CD3, CD4, CD8;
# macrophages, CD14, Aif1;

gene_immune <- c("Cd19","Ms4a1", 
                 "Cd3e","Cd3d","Cd3g","Cd3eap","Cd4","Cd8a","Cd8b1",
                 "Cd14","Aif1","Aif1l",
                 "Xcr1")
Idents(filtered_CSawyer_intact_Imm) <- filtered_CSawyer_intact_Imm$RNA_snn_res.3
AvgExpr <- AverageExpression(filtered_CSawyer_intact_Imm, features = gene_immune)
AvgExpr <- AvgExpr[[1]]
AvgExpr <- as.data.frame(AvgExpr)
AvgExpr <- t(AvgExpr)
AvgExpr <- round(AvgExpr, 3)

Idents(filtered_CSawyer_intact_Imm) <- filtered_CSawyer_intact_Imm$RNA_snn_res.3
VlnPlot(filtered_CSawyer_intact_Imm, features = gene_immune, stack = T, flip = T)

Idents(filtered_CSawyer_intact_Imm) <- "RNA_snn_res.3"
filtered_CSawyer_intact_Imm <-  RenameIdents(object = filtered_CSawyer_intact_Imm,
                                             "0" = "Macrophage",
                                             "1" = "T cell",
                                             "2" = "Macrophage",
                                             "3" = "Macrophage",
                                             "4" = "Macrophage",
                                             "5" = "Macrophage",
                                             "6" = "T cell",
                                             "7" = "Macrophage",
                                             "8" = "Macrophage",
                                             "9" = "Macrophage",
                                             "10" = "Macrophage",
                                             "11" = "Macrophage",
                                             "12" = "Macrophage",
                                             "13" = "Macrophage",
                                             "14" = "Dendritic cell",
                                             "15" = "B cell",
                                             "16" = "T cell",
                                             "17" = "Macrophage",
                                             "18" = "Macrophage",
                                             "19" = "Macrophage",
                                             "20" = "Macrophage",
                                             "21" = "Macrophage",
                                             "22" = "Macrophage"
)
filtered_CSawyer_intact_Imm$Immune_celltype <- Idents(filtered_CSawyer_intact_Imm)
p1 <- DimPlot(filtered_CSawyer_intact_Imm, group.by = "RNA_snn_res.3", label = T)
p2 <-DimPlot(filtered_CSawyer_intact_Imm, group.by = "Immune_celltype")
p3 <-DimPlot(filtered_CSawyer_intact_Imm, group.by = "predType")
plot_grid(p1, p2, p3, ncol = 2)

# save(filtered_CSawyer_intact_Imm, file = "RData/filtered_CSawyer_intact_Imm_XGBoost.RData")
```

#### xgboost for Intact CSawyers (immune only) ####

```{r}
#install.packages("xgboost")
library(xgboost)
library(tidymodels)

set.seed(123)
data<- filtered_CSawyer_intact_Imm@assays$RNA@scale.data
data<- data[rownames(data) %in% gene_immune,]
data<- t(data) %>% as.data.frame()
data$cell_type <- filtered_CSawyer_intact_Imm$Immune_celltype
data$cell_type<- factor(data$cell_type)
data$cell_barcode<- rownames(data)

data_split <- initial_split(data, strata = "cell_type")
data_train <- training(data_split)
data_test <- testing(data_split)

# 10 fold cross validation
data_fold <- vfold_cv(data_train, v = 10)

bt_recipe <- 
  recipe(formula = cell_type ~ ., data = data_train) %>%
  update_role(cell_barcode, new_role = "ID") %>%
  step_zv(all_predictors())

## feature importance sore to TRUE, tune mtry and number of trees
bt_spec <- 
  boost_tree(mtry = tune(), trees = tune()) %>%
  set_engine("xgboost", importance = TRUE) %>%
  set_mode("classification")

bt_workflow <- workflow() %>% 
  add_recipe(bt_recipe) %>% 
  add_model(bt_spec)

bt_grid<- grid_regular(mtry(range= c(1,20)), trees(), levels = 3)

# library(doParallel)
# doParallel::registerDoParallel()

# # save_pred = TRUE for later ROC curve
tune_res <- tune_grid(
  bt_workflow,
  resamples = data_fold, 
  grid = bt_grid,
  control = control_grid(save_pred = TRUE)
)

# check which penalty is the best 
autoplot(tune_res)
best_penalty <- select_best(tune_res, metric = "accuracy")
best_penalty

bt_final <- finalize_workflow(bt_workflow, best_penalty)
bt_final_fit_Imm <- fit(bt_final, data = data_train)

## confusion matrix, perfect classification! 
predict(bt_final_fit_Imm, new_data = data_test) %>%
  bind_cols(data_test %>% select(cell_type)) %>%
  conf_mat(truth = cell_type, estimate = .pred_class)

# save(bt_final_fit_Imm, file = "RData/XGBoost_CSawyer_T00int_Immune.RData")
```

#### 2c. Subset to Stromal only (filtered_CSawyer_intact) ####

```{r}

Idents(filtered_CSawyer_intact) <- filtered_CSawyer_intact$highlevelCelltype
filtered_CSawyer_intact_Str <- subset(filtered_CSawyer_intact, idents = "Stromal")
filtered_CSawyer_intact_Str <- CreateSeuratObject(counts = filtered_CSawyer_intact_Str@assays$RNA@counts,
                                                  meta.data = filtered_CSawyer_intact_Str@meta.data %>% select(batchID,highLevelPred,highLevelPredAmbig,predType, predTypeAmbig,
                                                                                                               sampleInfo,highlevelCelltype))

filtered_CSawyer_intact_Str <- NormalizeData(filtered_CSawyer_intact_Str, normalization.method = "LogNormalize", scale.factor = 10000)
filtered_CSawyer_intact_Str <- FindVariableFeatures(filtered_CSawyer_intact_Str, selection.method = "vst", nfeatures = 2000, verbose = T)
filtered_CSawyer_intact_Str <- ScaleData(filtered_CSawyer_intact_Str,  do.center = FALSE)
filtered_CSawyer_intact_Str <- RunOptimizeALS(filtered_CSawyer_intact_Str, k = 21, lambda = 5)
filtered_CSawyer_intact_Str <- RunQuantileNorm(filtered_CSawyer_intact_Str)
filtered_CSawyer_intact_Str <- FindNeighbors(filtered_CSawyer_intact_Str, reduction = "iNMF_raw", dims = 1:20)
filtered_CSawyer_intact_Str <- FindClusters(filtered_CSawyer_intact_Str, resolution = c(0.5, 0.8,1,3,5))
filtered_CSawyer_intact_Str <- RunUMAP(filtered_CSawyer_intact_Str, dims = 1:ncol(filtered_CSawyer_intact_Str[["iNMF_raw"]]), reduction = "iNMF_raw")
Idents(filtered_CSawyer_intact_Str) <- filtered_CSawyer_intact_Str$RNA_snn_res.0.5

DimPlot(filtered_CSawyer_intact_Str, group.by = "RNA_snn_res.0.5", label = T)

# Check cell type
# refer to Karthaus et al. Science 2020 (CSawyers) mouse clustering
# endothelium CD31, Vwf;
# lymphatic endothelium, CD31, Prox1; 
# glia, Sox10;
# myofibroblast, Acta2, Myh11, Rspo3;
# smooth muscle, Acta2, Notch3;
# mesenchymal 1, Col5a2, Lama2, Zeb1 Wnt2, Wnt6, Wnt10a, Rorb;
# mesenschymal 2,Col5a2, Lama2, Zeb1, Sult1e1, Fgf10, Rspo1;

gene_stromal <- c("Pecam1", "Vwf",
                  "Prox1", "Sox10",
                  "Acta2", "Myh11", "Rspo3","Notch3",
                  "Col5a2", "Lama2", "Zeb1", "Wnt2", "Wnt6", "Wnt10a", "Rorb",
                  "Sult1e1", "Fgf10","Rspo1")
Idents(filtered_CSawyer_intact_Str) <- filtered_CSawyer_intact_Str$RNA_snn_res.0.5
AvgExpr <- AverageExpression(filtered_CSawyer_intact_Str, features = gene_stromal)
AvgExpr <- AvgExpr[[1]]
AvgExpr <- as.data.frame(AvgExpr)
AvgExpr <- t(AvgExpr)
AvgExpr <- round(AvgExpr, 3)

Idents(filtered_CSawyer_intact_Str) <- filtered_CSawyer_intact_Str$RNA_snn_res.0.5
VlnPlot(filtered_CSawyer_intact_Str, features = gene_stromal, stack = T, flip = T)

Idents(filtered_CSawyer_intact_Str) <- "RNA_snn_res.0.5"
filtered_CSawyer_intact_Str <-  RenameIdents(object = filtered_CSawyer_intact_Str,
                                             "0" = "Mesenchymal 1",
                                             "1" = "Mesenchymal 1",
                                             "2" = "Mesenchymal 2",
                                             "3" = "Endothelial",
                                             "4" = "Endothelial",
                                             "5" = "Mesenchymal 2",
                                             "6" = "Myofibroblast/Smooth muscle",
                                             "7" = "Mesenchymal 2",
                                             "8" = "Myofibroblast/Smooth muscle",
                                             "9" = "Lymphatic endothelium",
                                             "10" = "Glial",
                                             "11" = "Mesenchymal 1",
                                             "12" = "Mesenchymal 1",
                                             "13" = "Myofibroblast/Smooth muscle"
)
filtered_CSawyer_intact_Str$stromal_celltype <- Idents(filtered_CSawyer_intact_Str)
p1 <- DimPlot(filtered_CSawyer_intact_Str, group.by = "RNA_snn_res.0.5", label = T)
p2 <-DimPlot(filtered_CSawyer_intact_Str, group.by = "stromal_celltype")
p3 <-DimPlot(filtered_CSawyer_intact_Str, group.by = "predType")
plot_grid(p1, p2, p3, ncol = 2)

# save(filtered_CSawyer_intact_Str, file = "RData/filtered_CSawyer_intact_Str_XGBoost.RData")

```

#### xgboost for Intact CSawyers (Stromal only) ####

```{r}
#install.packages("xgboost")
library(xgboost)
library(tidymodels)

set.seed(123)
data<- filtered_CSawyer_intact_Str@assays$RNA@scale.data
data<- data[rownames(data) %in% gene_stromal,]
data<- t(data) %>% as.data.frame()
data$cell_type <- filtered_CSawyer_intact_Str$stromal_celltype
data$cell_type<- factor(data$cell_type)
data$cell_barcode<- rownames(data)

data_split <- initial_split(data, strata = "cell_type")
data_train <- training(data_split)
data_test <- testing(data_split)

# 10 fold cross validation
data_fold <- vfold_cv(data_train, v = 10)

bt_recipe <- 
  recipe(formula = cell_type ~ ., data = data_train) %>%
  update_role(cell_barcode, new_role = "ID") %>%
  step_zv(all_predictors())

## feature importance sore to TRUE, tune mtry and number of trees
bt_spec <- 
  boost_tree(mtry = tune(), trees = tune()) %>%
  set_engine("xgboost", importance = TRUE) %>%
  set_mode("classification")

bt_workflow <- workflow() %>% 
  add_recipe(bt_recipe) %>% 
  add_model(bt_spec)

bt_grid<- grid_regular(mtry(range= c(1, 20)), trees(), levels = 3)

# library(doParallel)
# doParallel::registerDoParallel()

# # save_pred = TRUE for later ROC curve
tune_res <- tune_grid(
  bt_workflow,
  resamples = data_fold, 
  grid = bt_grid,
  control = control_grid(save_pred = TRUE)
)

# check which penalty is the best 
autoplot(tune_res)
best_penalty <- select_best(tune_res, metric = "accuracy")
best_penalty

bt_final <- finalize_workflow(bt_workflow, best_penalty)
bt_final_fit_Stromal <- fit(bt_final, data = data_train)

## confusion matrix, perfect classification! 
predict(bt_final_fit_Stromal, new_data = data_test) %>%
  bind_cols(data_test %>% select(cell_type)) %>%
  conf_mat(truth = cell_type, estimate = .pred_class)

# save(bt_final_fit_Stromal, file = "RData/XGBoost_CSawyer_T00int_Stromal.RData")
```

#### 3. Load all Karthaus CSawyers (intact and castrate) ####

```{r}
# load Karthaus CSawyers
CSawyer 
# Subset to Intact vs CastD28 only
Idents(CSawyer) <- CSawyer$sampleType
CSawyer_subset <- subset(CSawyer, 
                         idents = c("Intact","Cast") , 
                         invert = F)
Idents(CSawyer_subset) <- CSawyer_subset$batchID
CSawyer_subset <- subset(CSawyer_subset, idents = c("T00_Epi_1","T00_Epi_2",
                                                    "T00_NonEpi_1", "T00_NonEpi_2", "T00_Unsorted_1",
                                                    "T01_Cast_Day1_1" , "T01_Cast_Day1_2" , "T02_Cast_Day7_1",  "T02_Cast_Day7_2", 
                                                    "T03_Cast_Day14_1","T03_Cast_Day14_2"), 
                         invert = T)
table(CSawyer_subset$batchID)

# Compute percent mito ratio
CSawyer_subset$mitoRatio <- PercentageFeatureSet(object = CSawyer_subset, pattern = "^mt-")
CSawyer_subset$mitoRatio <- CSawyer_subset@meta.data$mitoRatio / 100

# filter
filtered_CSawyer_subset <- subset(
  x = CSawyer_subset,
  subset =  nFeature_RNA > 500 &
    nCount_RNA > 1000 &
    mitoRatio < 0.15)

table(filtered_CSawyer_subset$batchID)

# run basic Seurat to get the cluster
filtered_CSawyer_subset <- NormalizeData(filtered_CSawyer_subset, normalization.method = "LogNormalize", scale.factor = 10000)
filtered_CSawyer_subset <- FindVariableFeatures(filtered_CSawyer_subset, selection.method = "vst", nfeatures = 2000, verbose = T)
filtered_CSawyer_subset <- ScaleData(filtered_CSawyer_subset, split.by = "batchID", do.center = FALSE)
filtered_CSawyer_subset <- RunOptimizeALS(filtered_CSawyer_subset, k = 21, lambda = 5, split.by = "batchID")
filtered_CSawyer_subset <- RunQuantileNorm(filtered_CSawyer_subset, split.by = "batchID")
filtered_CSawyer_subset <- FindNeighbors(filtered_CSawyer_subset, reduction = "iNMF_raw", dims = 1:20)
filtered_CSawyer_subset <- FindClusters(filtered_CSawyer_subset, resolution = c(0.5, 0.8,1,3,5))
filtered_CSawyer_subset <- RunUMAP(filtered_CSawyer_subset, dims = 1:ncol(filtered_CSawyer_subset[["iNMF_raw"]]), reduction = "iNMF_raw")

## Split by high level lineage
Idents(filtered_CSawyer_subset) <- filtered_CSawyer_subset$RNA_snn_res.0.8
AvgExpr <- AverageExpression(filtered_CSawyer_subset, features = c("Epcam","Ptprc"), add.ident = "sampleInfo")
AvgExpr <- AvgExpr[[1]]
AvgExpr <- as.data.frame(AvgExpr)
AvgExpr <- t(AvgExpr)

Idents(filtered_CSawyer_subset) <- "RNA_snn_res.0.8"
VlnPlot(filtered_CSawyer_subset, features = c("Epcam","Ptprc"), stack = T, flip = T, split.by = "sampleInfo")

filtered_CSawyer_subset <-  RenameIdents(object = filtered_CSawyer_subset,
                                         "0" = "Epithelial",
                                         "1" = "Epithelial",
                                         "2" = "Epithelial",
                                         "3" = "Epithelial",
                                         "4" = "Stromal",
                                         "5" = "Stromal",
                                         "6" = "Epithelial",
                                         "7" = "Stromal",
                                         "8" = "Epithelial",
                                         "9" = "Immune",
                                         "10" = "Stromal",
                                         "11" = "Stromal",
                                         "12" = "Immune",
                                         "13" = "Epithelial",
                                         "14" = "Epithelial",
                                         "15" = "Immune",
                                         "16" = "Stromal",
                                         "17" = "Epithelial",
                                         "18" = "Stromal",
                                         "19" = "Immune",
                                         "20" = "Epithelial",
                                         "21" = "Stromal",
                                         "22" = "Immune",
                                         "23" = "Immune",
                                         "24" = "Stromal",
                                         "25" = "Stromal",
                                         "26" = "Immune"
                                         
)
filtered_CSawyer_subset$highlevelCelltype <- Idents(filtered_CSawyer_subset)
p1 <- DimPlot(filtered_CSawyer_subset, group.by = "highlevelCelltype", label = F, split.by = "sampleInfo")
p2 <- DimPlot(filtered_CSawyer_subset, group.by = "highLevelPred", label = F, split.by = "sampleInfo")
plot_grid(p1,p2, ncol = 1)

# save(filtered_CSawyer_subset, file = "RData/filtered_CSawyer_subset_highlevelCellType.RData")
```

#### 4a. subset to epithelial clusters only to get classifier ####

```{r}
Idents(filtered_CSawyer_subset) <- filtered_CSawyer_subset$highlevelCelltype
filtered_CSawyer_subset_Epi <- subset(filtered_CSawyer_subset, idents = "Epithelial")
table(filtered_CSawyer_subset_Epi$highlevelCelltype)

# predict via bt_fit
# load("RData/XGBoost_CSawyer_T00int_Epi_withbasalsubsets_final.RData")
set.seed(123)
data_CSawyer <- filtered_CSawyer_subset_Epi@assays$RNA@data
data_CSawyer <- data_CSawyer[rownames(data_CSawyer) %in% c(gene_epi),]
data_CSawyer<- t(data_CSawyer) %>% as.data.frame()
data_CSawyer$RNA_snn_res.1 <- filtered_CSawyer_subset_Epi$RNA_snn_res.1
data_CSawyer$RNA_snn_res.1 <- factor(data_CSawyer$RNA_snn_res.1)
data_CSawyer$cell_barcode<- rownames(data_CSawyer)
# colnames(data_CSawyer)[6] <- "Nkx3-1"

# Check that the column names match
all(colnames(data_CSawyer) == colnames(bt_final_fit_Epi))
# Make predictions
pred_CSawyer_Epi <- predict(bt_final_fit_Epi , new_data = data_CSawyer) %>%
  bind_cols(data_CSawyer %>% select(RNA_snn_res.1, cell_barcode)) 

filtered_CSawyer_subset_Epi$pred_celltype <- ifelse(colnames(filtered_CSawyer_subset_Epi) %in% pred_CSawyer_Epi$cell_barcode, 
                                                    as.character(pred_CSawyer_Epi$.pred_class), "NA")
DimPlot(filtered_CSawyer_subset_Epi, group.by = "pred_celltype", split.by = "sampleInfo")
DimPlot(filtered_CSawyer_subset_Epi, group.by = "predType", split.by = "sampleInfo")


# save(filtered_CSawyer_subset_Epi, file="RData/filtered_CSawyer_subset_Epi_XGBoost.RData")

```

#### 4b. subset to Immune clusters only to get classifier ####


```{r}

Idents(filtered_CSawyer_subset) <- filtered_CSawyer_subset$highlevelCelltype
filtered_CSawyer_subset_Imm <- subset(filtered_CSawyer_subset, idents = "Immune")
table(filtered_CSawyer_subset_Imm$highlevelCelltype)

# predict via bt_fit
# load("RData/XGBoost_CSawyer_T00int_Immune.RData")
set.seed(123)
data_CSawyer <- filtered_CSawyer_subset_Imm@assays$RNA@data
data_CSawyer <- data_CSawyer[rownames(data_CSawyer) %in% c(gene_immune),]
data_CSawyer<- t(data_CSawyer) %>% as.data.frame()
data_CSawyer$RNA_snn_res.1 <- filtered_CSawyer_subset_Imm$RNA_snn_res.1
data_CSawyer$RNA_snn_res.1 <- factor(data_CSawyer$RNA_snn_res.1)
data_CSawyer$cell_barcode<- rownames(data_CSawyer)
# colnames(data_CSawyer)[6] <- "Nkx3-1"

# Check that the column names match
all(colnames(data_CSawyer) == colnames(bt_final_fit_Imm))
# Make predictions
pred_CSawyer_Epi <- predict(bt_final_fit_Imm , new_data = data_CSawyer) %>%
  bind_cols(data_CSawyer %>% select(RNA_snn_res.1, cell_barcode)) 

filtered_CSawyer_subset_Imm$pred_celltype <- ifelse(colnames(filtered_CSawyer_subset_Imm) %in% pred_CSawyer_Epi$cell_barcode, 
                                                    as.character(pred_CSawyer_Epi$.pred_class), "NA")

Idents(filtered_CSawyer_subset_Imm) <- filtered_CSawyer_subset_Imm$pred_celltype
Marker_filtered_CSwayer_subset_Imm <- FindMarkers(filtered_CSawyer_subset_Imm, only.pos = F, ident.1 = "T04_Cast_Day28",
                                                  group.by = "sampleInfo", subset.ident = "B cell")
Marker_filtered_CSwayer_subset_Imm <- Marker_filtered_CSwayer_subset_Imm %>% filter(p_val_adj < 0.05)

# png(filename = "Output/For_ARdel_project/Dimplot_filtered_CSawyer_subset_Imm.# png", width = 1300, height = 1000, res = 200)
DimPlot(filtered_CSawyer_subset_Imm, group.by = "pred_celltype", split.by = "sampleInfo")
# dev.off()

# save(filtered_CSawyer_subset_Imm, file="RData/filtered_CSawyer_subset_Imm_XGBoost.RData")
```

#### 4c. subset to Stromal clusters only to get classifier ####

```{r}
Idents(filtered_CSawyer_subset) <- filtered_CSawyer_subset$highlevelCelltype
filtered_CSawyer_subset_Str <- subset(filtered_CSawyer_subset, idents = "Stromal")
table(filtered_CSawyer_subset_Str$highlevelCelltype)

# predict via bt_fit
# load("RData/XGBoost_CSawyer_T00int_Stromal.RData")
set.seed(123)
data_CSawyer <- filtered_CSawyer_subset_Str@assays$RNA@data
data_CSawyer <- data_CSawyer[rownames(data_CSawyer) %in% c(gene_stromal),]
data_CSawyer<- t(data_CSawyer) %>% as.data.frame()
data_CSawyer$RNA_snn_res.1 <- filtered_CSawyer_subset_Str$RNA_snn_res.1
data_CSawyer$RNA_snn_res.1 <- factor(data_CSawyer$RNA_snn_res.1)
data_CSawyer$cell_barcode<- rownames(data_CSawyer)
# colnames(data_CSawyer)[6] <- "Nkx3-1"

# Check that the column names match
all(colnames(data_CSawyer) == colnames(bt_final_fit_Stromal))
# Make predictions
pred_CSawyer_Epi <- predict(bt_final_fit_Stromal , new_data = data_CSawyer) %>%
  bind_cols(data_CSawyer %>% select(RNA_snn_res.1, cell_barcode)) 

filtered_CSawyer_subset_Str$pred_celltype <- ifelse(colnames(filtered_CSawyer_subset_Str) %in% pred_CSawyer_Epi$cell_barcode, 
                                                    as.character(pred_CSawyer_Epi$.pred_class), "NA")

# png(filename = "Output/For_ARdel_project/Dimplot_filtered_CSawyer_subset_Str.# png", width = 1300, height = 1000, res = 200)
DimPlot(filtered_CSawyer_subset_Str, group.by = "pred_celltype", split.by = "sampleInfo")
# dev.off()

# save(filtered_CSawyer_subset_Str, file="RData/filtered_CSawyer_subset_Str_XGBoost.RData")

```
